Implementing Reactive Systems Directly\\with I/O Automata in ioa++

### 2012-02-24 papers/paper2/case_study.tex
Down to 17 pages.

In this section, we illustrate how ioa++ can be used to build real distributed systems.
In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
The intermediate layers contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.

@@

Nodes in a distributed system may be arranged in a ring so that two consecutive nodes are connected by a directed channel.
The goal is to designate one of the nodes as the leader of the ring.
The solution we adopt is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
The LCR algorithm assumes that every node has a unique identifier (UID).
When the protocol begins, each node sends its UID to its successor.
When a node receives a UID from its predecessor that is greater than its own, it forwards the UID to its successor.
When a node receives its own UID, it elects itself the leader of the ring.
We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}} of the ioa++ package.

A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
The tcp\_ring\_automaton listed in examples/tcp\_ring\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{constellations} to manage the dynamic constellation of automata and recover from errors.
The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.

\begin{figure}
\center
\includegraphics[width=\columnwidth]{tcp_lcr_automaton}
\caption{Architecture of the TCP LCR system.}
\label{tcp_lcr_automaton}
\end{figure}

The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.

Figure~\ref{tcp_lcr_automaton} shows the resulting layered architecture of the TCP LCR system.
The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the application logic.
As this simple but representative example illustrates, the ioa++ framework readily supports the development of architecturally layered component-based distributed systems involving non-trivial dynamics.

>>>>

In this section, we illustrate how ioa++ can be used to build real distributed systems.
In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
Intermediate layers offer generic coordination mechanisms that serve to abstract away the low-level details from the application logic.

@@

Nodes in a distributed system may be arranged in a ring with one of the nodes as the leader of the ring, e.g., elected per the asynchronous LCR algorithm~\cite{lynch1996distributed}.
The LCR algorithm assumes that every node has a unique identifier (UID).
When the protocol begins, each node sends its UID to its successor.
When a node receives a UID from its predecessor that is greater than its own, it forwards the UID to its successor.
When a node receives its own UID, it elects itself the leader of the ring.
We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
Our implementation of the asynchronous LCR automaton is also freely available as open-source\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}}.

A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
The tcp\_ring\_automaton listed in examples/tcp\_ring\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{programming_model} to manage the dynamic constellation of automata and recover from errors.
The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.

%% \begin{figure}
%% \center
%% \includegraphics[width=\columnwidth]{tcp_lcr_automaton}
%% \caption{Architecture of the TCP LCR system.}
%% \label{tcp_lcr_automaton}
%% \end{figure}

The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
In the resulting layered architecture of the TCP LCR system, the bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the application logic.
As this simple but representative example illustrates, the ioa++ framework readily supports the development of architecturally layered component-based distributed systems involving non-trivial dynamics.

### 2012-02-24 papers/paper2/component_model.tex
Down to 17 pages.

\subsection{Background: I/O Automata}

The Input/Output (I/O) automata model was developed to model reactive concurrent and asynchronous systems~\cite{lynch1996distributed}.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action only changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\ifjournal
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.
\fi
\ifjournal
Often, a local action consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected for execution.
The effect is a function that computes the next state of the automaton from the current state.
An input action presented in this style consists solely of an effect.
\fi
I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effects of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action (an output or internal action~\cite{lynch1996distributed}) and then if the precondition is true applying the effect.
The scheduler is assumed to be \emph{fair,} meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.

\subsection{Component Structure\label{component_structure}}

@@

\paragraph*{Motivation for I/O automata based components.}
%The I/O automata formalism was developed to model reactive systems~\cite{lynch1987hierarchical} and has been used in the design and verification of concurrent and distributed systems.
I/O automata are a good foundation for such components because they have independent state, straightforward interfaces, and precise semantics for event generation, distribution, and consumption resulting in well-defined interactions under composition.
\emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1994atomic} which allows a \emph{parent} to contain zero or more \emph{children}.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

@@

\paragraph*{Explicit composition.}
Composition in the I/O automata model is defined at the automaton level and is based on actions having the same name.
To prepare for dynamic composition, we redefine composition at the action level.
An explicit association between an output action and input action is called a \emph{binding}.
%% The automaton that prescribes a binding is said to \emph{own} the binding.
%% A binding is denoted $(output, input, owner)$.
The set of bindings in a system must adhere to four essential rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
Second, an input action can be bound to at most one output action.
Third, an output action cannot be bound to more than one input action in the same automaton.
Fourth, an output action cannot be bound to an input action in the same automaton.
The second, third, and fourth rules exist because they admit ambiguity into the relationship between the states of the associated automata.
Graphically, an arrow originating in one component automaton must terminate in another component automaton as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph*{Concurrent execution.}
Recall that composing I/O automata consists of concatenating state variables and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $out!(s)$ and $process1$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

\subsection{Dynamics\label{dynamics}}

>>>>

\paragraph*{I/O automata.}
The Input/Output (I/O) automata model was developed to model reactive concurrent and asynchronous systems~\cite{lynch1996distributed}.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action only changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\ifjournal
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.
\fi
\ifjournal
Often, a local action consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected for execution.
The effect is a function that computes the next state of the automaton from the current state.
An input action presented in this style consists solely of an effect.
\fi
I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effects of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action (an output or internal action~\cite{lynch1996distributed}) and then if the precondition is true applying the effect.
The scheduler is assumed to be \emph{fair,} meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.

@@

\paragraph*{Motivation for I/O automata based components.}
%The I/O automata formalism was developed to model reactive systems~\cite{lynch1987hierarchical} and has been used in the design and verification of concurrent and distributed systems.
I/O automata are a good foundation for such components because they have independent state, straightforward interfaces, and precise semantics for event generation, distribution, and consumption resulting in well-defined interactions under composition.
\emph{Hierarchical decomposition} is an established technique for I/O automata~\cite{lynch1994atomic} which allows a \emph{parent} to contain zero or more \emph{children}.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

@@

\paragraph*{Explicit composition.}
Composition of I/O automata is based on actions having the same name.
To prepare for dynamic composition, we redefine composition at the action level.
An explicit association between an output action and input action is called a \emph{binding}.
%% The automaton that prescribes a binding is said to \emph{own} the binding.
%% A binding is denoted $(output, input, owner)$.
The set of bindings in a system must adhere to four essential rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
Second, an input action can be bound to at most one output action.
Third, an output action cannot be bound to more than one input action in the same automaton.
Fourth, an output action cannot be bound to an input action in the same automaton.
The second, third, and fourth rules exist because they admit ambiguity into the relationship between the states of the associated automata.
Graphically, an arrow originating in one component automaton must terminate in another component automaton as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph*{Concurrent execution.}
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $out!(s)$ and $process1$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

### 2012-02-24 papers/paper2/design.tex
Down to 17 pages.

The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} for POSIX environments.
The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.

Conceptually, the set of automata $A$ and set of bindings $B$ belong to the system automaton.
However, the state they encode is shared and is used in the execution of every action.
A \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
The opportunities for concurrent execution discussed in Section~\ref{component_structure} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
Binding predicates do not modify $B$ and therefore do not affect concurrent execution.
The scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.

@@

\subsection{Scheduler\label{scheduling}}

The scheduler is responsible for both selecting and executing local actions.
These two activities are decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
Input actions are executed with corresponding output actions, according to the current set of bindings.
The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
The controller encodes a \emph{policy} that selects the next local action.
For the framework to be a valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
The run-time system can be initialized with different controllers to achieve different performance objectives.
The ioa++ framework distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.

\paragraph*{Explicit dynamic scheduling.}
In our framework implementation, automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
%The act of declaring an action to the scheduler is called \emph{scheduling}.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare actions to the scheduler (i.e., to \emph{schedule} them) after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.
Scheduling is idempotent to prevent unbounded growth of scheduler data structures.

\paragraph*{Binding predicates and scheduling.}
%% This assumption is not implicit because we state that the preconditions are independent.
The preceding discussion on scheduling assumed that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind action changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after each successful bind or unbind.

\paragraph*{Delayed actions.}
Scheduling activities for a future time is a useful feature for many systems because it allows the system to sleep between activities. For example, the framework should support the development of automata that act as timers and alarms.
Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph*{File descriptor events.}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects may adversely affect the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is asynchronously ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
The ioa++ framework uses the Reactor pattern since (1) it is simpler and (2) many operating systems have non-blocking I/O and some kind of synchronous I/O (de)multiplexing, whereas asynchronous I/O is less widespread.

Our use of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
A user automaton creates a file descriptor and configures it for non-blocking I/O.
The automaton then tells the scheduler to release an action (typically an internal action) whenever the automaton becomes ready for reading (or writing).
Subsequent requests made using the same file descriptor and event (ready for reading, ready for writing) are ignored.
Once the file descriptor is ready, the action is released for consideration by the controller.
The scheduler also provides a special function for closing file descriptors that purges them from the reactor.

\subsection{System Actions\label{system_action_section}}

Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with an indication of either success or failure.
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple constellations but becomes unwieldy when creating more complicated ones.
Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.

>>>>

The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} for POSIX environments.
The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
Conceptually, the set of automata $A$ and set of bindings $B$ belong to the system automaton.
However, the state they encode is shared and is used in the execution of every action.
The \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_model} and the system automaton.
The opportunities for concurrent execution discussed in Section~\ref{component_model} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
Binding predicates do not modify $B$ and therefore do not affect concurrent execution.
The scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.

@@

The scheduler is responsible for both selecting and executing local actions.
These two activities are decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
Input actions are executed with corresponding output actions, according to the current set of bindings.
The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
The controller encodes a \emph{policy} that selects the next local action.
For fidelity to the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
The ioa++ distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.

\paragraph*{Explicit dynamic scheduling.}
Automata declare \emph{dynamically} the actions they would like the scheduler to consider, and executing an action may change the state of the automaton which in turn may cause the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare new actions to the scheduler after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.
Scheduling is idempotent to prevent unbounded growth of scheduler data structures.

\paragraph*{Binding predicates and scheduling.}
%% This assumption is not implicit because we state that the preconditions are independent.
In a static system, executing an action in one automaton cannot change a precondition in another automaton.
In a dynamic system, executing an action in the system automaton can change preconditions in other automata through binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind action changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
This example is representative of situations where an automaton is waiting for one or more of its outputs to be bound.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after each successful bind or unbind.

\paragraph*{Delayed actions.}
Scheduling activities for a future time may be useful to allow the system to sleep between activities.
For example, the framework should support automata that act as timers and alarms.
Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph*{File descriptor events.}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects may adversely affect the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is asynchronously ready for I/O and (2) to perform non-blocking I/O operations.
This causes an action to be scheduled when a file descriptor is ready for reading or writing.
A user automaton creates a file descriptor and configures it for non-blocking I/O.
The automaton then tells the scheduler to release a local action whenever the automaton becomes ready for reading or writing.
Subsequent requests made using the same file descriptor and event are ignored.
Once the file descriptor is ready, the action is released for consideration by the controller.

\paragraph*{System actions.}
Our approach to system actions follows the model introduced in Section~\ref{component_model}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with an indication of either success or failure.
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple constellations but becomes unwieldy when creating more complicated ones.
Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.

### 2012-02-24 papers/paper2/evaluation.tex
Removing material.
Down to 17 pages.

For this experiment, we used a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{scheduling}.
When the scheduler is configured to use a single thread, all actions are executed by that same thread.
When two threads are used, one thread executes the actions of one $R$ automaton while the other thread executes the actions of the other $R$ automaton.

@@

\begin{figure}
\center
\includegraphics[width=\columnwidth]{speed_up}
\caption{Speed-up for system $S$.\label{speed_up}}
\end{figure}

Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with two threads versus one thread.
Confidence intervals are shown but are negligible with the largest being a speed-up range of $\pm$0.0165.
When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
Consequently, every action must be serialized yielding a maximum speed-up of 1.
When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.

The speed-up shows a strong dependence on the duration of each local action which is proportional to $N$.
A small value for $N$ means that very little time is spent executing automaton code.
Consequently, a greater fraction of time is spent executing framework code which includes various synchronization calls to the pthreads library.
For a small enough $N$, this overhead dominates the execution time.
This situation is exacerbated by multiple threads since they will actively interfere with one another which can be seen in the slow-down for small values of $N$.
Conversely, when $N$ is large, relatively little time is spent executing framework and synchronization code.
Consequently, contention is reduced allowing for greater speed-ups as indicated by Figure~\ref{speed_up}.

\begin{figure}
\center
\includegraphics[width=\columnwidth]{overhead1}
\caption{Overhead for system $S$ when using one thread.\label{overhead1}}
\end{figure}

\begin{figure}
\center
\includegraphics[width=\columnwidth]{overhead2}
\caption{Overhead for system $S$ when using two threads.\label{overhead2}}
\end{figure}

Figures~\ref{overhead1} and \ref{overhead2} show the average per thread of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions; confidence intervals are again negligible with largest being $\pm$0.002155 for Figure~\ref{overhead1} and $\pm$0.001781 for Figure~\ref{overhead2}.
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes from 0 to 1 because the system transitions from always acquiring two locks to always acquiring one lock.
The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends very little time on synchronization when $\rho = 1$.

>>>>

For this experiment, we used a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{design}.
When the scheduler is configured to use a single thread, all actions are executed by that same thread.
When two threads are used, one thread executes the actions of one $R$ automaton while the other thread executes the actions of the other $R$ automaton.

@@

Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with two threads versus one thread.
Confidence intervals are shown but are negligible with the largest being a speed-up range of $\pm$0.0165.
When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
Consequently, every action must be serialized yielding a maximum speed-up of 1.
When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.
The speed-up shows a strong dependence on the duration of each local action which is proportional to $N$.
A small value for $N$ means that very little time is spent executing automaton code.
Consequently, a greater fraction of time is spent executing framework code which includes various synchronization calls to the pthreads library.
For a small enough $N$, this overhead dominates the execution time.
This situation is exacerbated by multiple threads since they will actively interfere with one another which can be seen in the slow-down for small values of $N$.
Conversely, when $N$ is large, relatively little time is spent executing framework and synchronization code.
Consequently, contention is reduced allowing for greater speed-ups as indicated by Figure~\ref{speed_up}.

\begin{figure}
\center
\includegraphics[width=.72\columnwidth]{speed_up}
\caption{Speed-up for system $S$.\label{speed_up}}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.72\columnwidth]{overhead1}
\caption{Overhead for system $S$ when using one thread.\label{overhead1}}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.72\columnwidth]{overhead2}
\caption{Overhead for system $S$ when using two threads.\label{overhead2}}
\end{figure}

Figures~\ref{overhead1} and \ref{overhead2} show the average per thread of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions; confidence intervals are again negligible with largest being $\pm$0.002155 for Figure~\ref{overhead1} and $\pm$0.001781 for Figure~\ref{overhead2}.
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes from 0 to 1 because the system transitions from always acquiring two locks to always acquiring one lock.
The multi-threaded execution shows that the system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends very little time on synchronization when $\rho = 1$.
